---
title: "Exploring Trends in Shark Landings by US Commercial Fisheries"
echo: FALSE
messages: FALSE
---

# Data Description

**Data Source.**  
My data comes from the [NOAA Fisheries Commercial Landings Database](https://www.fisheries.noaa.gov/foss/f?p=215:200:1455472432287:::::), which tracks reported commercial landings of fishes. I utilized their website to select data for 12 large pelagic shark species landed across the U.S.


**Variables and Data Types.**  
My data contains commercial fisheries landings from 1915 to 2023, and provides insight into trends in shark landings by state.
The dataset includes the following variables:
 - Year, year of the recorded landings.
 -State, U.S. state 
 - NMFS Name, National Marine Fisheries Service common name of each species
 - Pounds and Metric Tons, weight of sharks landed
 - Dollars, market value of the landed sharks
 - Confidentiality, whether or not the data is  confidential
 - Collection, whether it was commercially or recreational caught
 - Scientific Name, scientific name of each species
 - Tsn,  Taxonomic Serial Numbers are a numeric code for the species
 - Source,original source of the data

**Research Questions**  

1. Are there trends in the value of shark landings (in dollars) over time?

2. Which shark species have the highest commercial landing weights (weight of all sharks cought) over time?

3. How do the prominent species landed vary by US state and region?

4. What data source (fisheries management organization) had the highest average shark landings in the last 5 years?

# Data Cleaning

To get my dataset ready for visualizations, I first cleaned the variable names to make them easier to work with, and viewed a summary of the data to identify what needed cleaning.
Then, I adjusted how certain pieces of information were stored: I made sure the year was treated as a number, removed symbols and commas from the weight and dollar columns so they could be calculated properly, and grouped together categories like state names and species names. I also removed rows that were marked “Confidential” since they didn’t include usable data, and deleted columns that weren’t relevant to my project. For any dollar or weight values that were recorded as zero, I marked them as missing so they wouldn’t affect the results. After this data cleaning, I saved the updated dataset and created a new dataset that showed the total pounds of sharks landed and total dollar value for each species, by state and year. I added a new column that grouped states into larger regions, such as “New England” or “Pacific Coast”—based on NOAA’s region definitions, and saved this dataset. I then created another dataset that totaled shark landings and dollar values by year alone. Finally, I created a wide-format version of the data, where each species had its own column, which may be helpful for certain and analysis. 

```{r packages, include=FALSE}
library(tidyverse)
library(janitor)
```


```{r load_and_explore_data, include=FALSE}
landings<-read.csv("data/FOSS_landings-2.csv",  skip = 1) #skip a row to make column names
landings<-clean_names(landings) #cleaning variable names
summary(landings) #checking variable types

```

```{r data_cleaning}
#transform the data to the correct variable types
landings_clean <- landings %>%
  mutate(
    year = as.numeric(year),
    pounds = parse_number(pounds),
    metric_tons = parse_number(metric_tons), #parse number removes commas and saves a numeric variable
    dollars = parse_number(dollars),
    state = as.factor(state), #saving as factor for grouping and categorical visualizations later on
    nmfs_name = as.factor(nmfs_name),
    scientific_name = as.factor(scientific_name),
    confidentiality = as.character(confidentiality),
    source = as.factor(source))
    

#remove rows I will not be using
landings_filtered <- landings_clean %>%
  filter(confidentiality != "Confidential")%>% #removing all rows that are confidential, because there is no data
  select(-tsn, -collection, -confidentiality)
  

#make sure that 0s are not counted as NAs, because there may be categories where $ needs to be $0, and we do nnot want to accidentally filter out meaningful data
landings_filtered <- landings_filtered %>%
  mutate(
    dollars = ifelse(dollars == 0, NA, dollars),
    pounds = ifelse(pounds == 0, NA, pounds),
    metric_tons = ifelse(metric_tons == 0, NA, metric_tons)
  )


# save to the data folder
write.csv(landings_filtered, file.path("data", "landings_filtered.csv"))
```

```{r create_landings_datasets}

# make a dataframe of landings by year and state, and species
landings_state1 <- landings_filtered %>%
  group_by(year, state, nmfs_name) %>%
  summarize(
    total_pounds = sum(pounds, na.rm = TRUE), #create total pounds of sharks landed per year, state, and species
    total_dollars = sum(dollars, na.rm = TRUE),  #create total dollars of sharks landed per year, state, and species
    .groups = "drop"
  )

# specifying which states belong to each region (as defined by NOAA), and adding it as a column
landings_state <- landings_state1 %>%
  mutate(region = case_when(
    state %in% c("MAINE", "NEW HAMPSHIRE", "VERMONT", "MASSACHUSETTS", "RHODE ISLAND", "CONNECTICUT") ~ "New England",
    state %in% c("DELAWARE", "MARYLAND", "VIRGINIA", "WEST VIRGINIA") ~ "Middle Atlantic",
    state %in% c("NORTH CAROLINA", "SOUTH CAROLINA", "GEORGIA", "FLORIDA") ~ "South Atlantic",
    state %in% c("TEXAS", "LOUISIANA", "MISSISSIPPI", "ALABAMA") ~ "Gulf",
    state %in% c("MICHIGAN", "OHIO", "INDIANA", "ILLINOIS", "WISCONSIN", "MINNESOTA", "NEW YORK") ~ "Great Lakes",
    state == "HAWAII" ~ "Hawaii",
    state %in% c("CALIFORNIA", "OREGON", "WASHINGTON") ~ "Pacific Coast",
    state %in% c("AMERICAN SAMOA", "GUAM", "NORTHERN MARIANA ISLANDS", "PACIFIC ISLANDS") ~ "Pacific Island Regions",
    state == "ALASKA" ~ "Alaska",
    state == "NEW JERSEY" ~ "Middle Atlantic",   
    state == "FLORIDA-EAST" ~ "South Atlantic",  
    state == "FLORIDA-WEST" ~ "South Atlantic", 
    state == "PROCESS AT SEA" ~ "PROCESS AT SEA",       
    TRUE ~ "Other"  # a default for any states not matching
  ))

# save to the data folder
write.csv(landings_state, file.path("data", "landings_state.csv"))


# make a dataframe of total landings by year and state
total_landings_year <- landings_state %>%
  group_by(year) %>%
  summarize(total_pounds = sum(total_pounds, na.rm = TRUE), 
            total_dollars = sum(total_dollars, na.rm = TRUE))

# save to the data folder
write.csv(total_landings_year, file.path("data", "total_landings_year.csv"))
```

> I used help from AI (ChatGPT) to create the regions dataframe, and to make sure the tideous process of spelling all state names and regions was done correctly.

```{r wide_data_format}
#in case I want to run any statistical analysis on the data, a wide formatted dataset may be useful
wide_landings_state <- landings_state %>%
  group_by(year, state, nmfs_name) %>% #group by year, state, and species
  summarise(
    total_pounds = sum(total_pounds, na.rm = TRUE),
    total_dollars = sum(total_dollars, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = nmfs_name,
    values_from = c(total_pounds, total_dollars)
  )

# save to the data folder
write.csv(wide_landings_state, file.path("data", "wide_landings_state.csv"))
```

# Data Visualization 1

This line graph shows total U.S. commercial shark landings from 1950 to 2020. The x-axis represents years in ten-year intervals, while ther are two lines on the y-axis. One line, in  blue, shows the total pounds of sharks landed, steadily increasing through the 1980s and peaking in the early 1990s before declining. The other line, in dark green, shows the total dollar value of those landings, which generally follows the same trend as total catches. The y-axes on both sides of the plot display pounds and dollars, respectively. As we can see, the most commercial shark landings occurred around the late 1990s ( coinciding with the highest dollar value of landings), with a secondary peak during the 2010s. The number of shark landings must be monitored, as there is high variability in recent years.


```{r first_plot_draft}
#| fig-cap: "Total pounds of U.S. commercial shark landings (blue), and total dollars (green) over time. Source: NOAA Fisheries."
#| fig-alt: "A line graph showing total U.S. commercial shark landings from 1950 to 2020. The x-axis represents years in ten-year intervals. One line, in  blue, shows the total pounds of sharks landed, steadily increasing through the 1980s and peaking in the early 1990s before declining. The other line, in dark green, shows the total dollar value of those landings, which generally follows the same trend as total catches. The y-axes on both sides of the plot display pounds and dollars, respectively."

# line plot of total commercial shark landings and their value over time
total_landings_year %>%
  ggplot() +
  geom_line(aes(x = year, y = total_pounds), color = "steelblue") + # plot total pounds on the y axis and year on the x axis
  geom_line(aes(x = year, y = total_dollars), color = "darkgreen") +  # plot second line with total dollars on the y axis and year on the x axis
  scale_x_continuous(
    breaks = seq(min(total_landings_year$year), max(total_landings_year$year), by = 10)) +  # set x axis breaks by 10 year increments
 scale_y_continuous(
    name = "Total Pounds",  
    breaks = seq(0, max(total_landings_year$total_pounds), by = 10000000),  # set y axis breaks every 10 million pounds
    labels = function(x) format(x, big.mark = ",", scientific = FALSE),  # add commas to y axis labels 
    sec.axis = sec_axis(~ ., name = "Total Dollars",   #add second y axis label 
                        labels = function(x) paste0("$", format(x, big.mark = ",", scientific = FALSE)))  # add commas and $ to y axis labels 
  ) +
  labs(
    title = "Total US Commercial Shark Landings Over Time",  # title the plot
    x = "Year",  # x axis label
    y = "Total Pounds",  # y axis label
    caption = "Source: NOAA Fisheries landings data") +  # always cite your data!
  theme_minimal()  # a clean plot theme that I'll probably change later


```

# Data Visualization 2

This line graph shows total commercial shark landings in pounds from 1950 to 2020 for the five most frequently caught species. Each species is a different color, layered to show how their relative contributions change over time. The predominantly caught shark over time is the Spiny Dogfish.

```{r}
#| label: top-species-over-time
#| fig-cap: "Top 5 commercially landed shark species over time."
#| fig-alt: "An area chart showing total commercial shark landings in pounds from 1950 to 2020 for the five most frequently caught species. Each species is a different color, layered to show how their relative contributions change over time. The predominantly caught shark over time is the Spiny Dogfish."

landings_filtered %>%
  group_by(nmfs_name) %>%
  summarise(total_pounds_all = sum(pounds, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(total_pounds_all)) %>%
  slice_head(n = 5) %>%
  pull(nmfs_name) -> top_species

landings_filtered %>%
  filter(nmfs_name %in% top_species) %>%
  group_by(year, nmfs_name) %>%
  summarise(total_pounds = sum(pounds, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = year, y = total_pounds, color = nmfs_name)) +
  geom_line() +
  labs(
    title = "Top 5 Shark Species by Commercial Landing Weight Over Time",
    x = "Year",
    y = "Total Pounds",
    color = "Species",
    caption = "Source: NOAA Fisheries landings data"
  ) +
  theme_minimal()


```

# Data Visualization 3

This horizontal bar chart shows U.S. states on the vertical axis and total pounds of sharks landed on the horizontal axis. Each bar represents the shark species with the highest commercial landing weight in that state. Bars are color-coded by shark species, and the total pounds landed are labeled at the end of each bar. The chart highlights the variation in dominant caught shark species across states based on commercial landing weight in pounds. The predominantly caught shark across states is the Spiny Dogfish.

```{r}
#| label: top-species-per-state
#| fig-cap: "Top shark species by total pounds landed per state by commercial fisheries."
#| fig-alt: "A horizontal bar chart showing U.S. states on the vertical axis and total pounds of sharks landed on the horizontal axis. Each bar represents the shark species with the highest commercial landing weight in that state. Bars are color-coded by shark species, and the total pounds landed are labeled at the end of each bar. The chart highlights the variation in dominant caught shark species across states based on commercial landing weight in pounds. The predominantly caught shark across states is the Spiny Dogfish, with Massachusets hving the highest total pounds landed."

top_species_per_state <- landings_state %>%
  group_by(state, nmfs_name) %>%
  summarise(total_pounds = sum(total_pounds, na.rm = TRUE), .groups = "drop") %>%
  group_by(state) %>%
  slice_max(total_pounds, n = 1) %>%  # get the top species per state
  ungroup()

ggplot(top_species_per_state, aes(x = reorder(state, total_pounds), y = total_pounds, fill = nmfs_name)) +
  geom_col(show.legend = TRUE) +
  geom_text(aes(label = scales::comma(total_pounds)), 
            hjust = -0.1, size = 3.5) +  # add pounds labels just outside the bars
  coord_flip() +
  labs(
    title = "Top Shark Species by Total Pounds Landed per State",
    x = "State",
    y = "Total Pounds",
    fill = "Shark Species"
  ) +
  theme_minimal() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)))  # add some space on the right for labels


```
# Data Visualization 4

This heatmap shows the total shark landings in pounds by U.S. coastal region from 1950 to 2022. Each tile represents a year-region combination, shaded by total pounds landed. Lighter shades indicate higher landings. The Pacific coast shows the earliest landings, while New England region shows highest landings (in 1990).

```{r}
#| label: "Shark-Landings-by-Region"
#| fig-cap: "Heatmap displaying annual shark landings (in pounds) across U.S. fisheries regions from 1950 to 2022. Darker colors indicate higher landings."
#| fig-alt: "A heatmap showing total shark landings in pounds by U.S. coastal region from 1950 to 2022. Each tile represents a year-region combination, shaded by total pounds landed. Lighter shades indicate higher landings. The Pacific coast shows the earliest landings, while New England region shows highest landings (in 1990)."

landings_state %>%
  group_by(region, year) %>%
  summarise(total_pounds = sum(total_pounds, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = year, y = fct_rev(region), fill = total_pounds)) +
  geom_tile() +
  scale_fill_viridis_c(labels = scales::comma) +
  labs(
    x = "Year",
    y = "Region",
    fill = "Total Pounds",
    title = "Heatmap of Shark Landings by Region and Year"
  ) +
  theme_minimal(base_size = 13)


```


# Data Visualization 5

This bar chart displays the average shark landing weights, measured in pounds, from various fisheries management organizations over the most recent five years (2015-2020). The Atlantic Coastal Cooperative Statistics Program has the highest average landings by a large margin (just over 1,600,000 lbs). The Pacific Fisheries Information Network, Exceptions, and the Alaska Fisheries Information Network are significantly lower than the Atlantic Coastal Cooperative Statistics Program. The chart provides a clear comparison of how shark landings differ across these organizations.

```{r}
#| label: ave-landings-source
#| fig-cap: "Bar chart of average shark landing weights (in pounds) by commercial fisheries over the last 5 years."
#| fig-alt: "This bar chart displays the average shark landing weights, measured in pounds, from various fisheries management organizations over the most recent five years (2015-2020). The Atlantic Coastal Cooperative Statistics Program has the highest average landings by a large margin (just over 1,600,000 lbs). The Pacific Fisheries Information Network, Exceptions, and the Alaska Fisheries Information Network are significantly lower than the Atlantic Coastal Cooperative Statistics Program. The chart provides a clear comparison of how shark landings differ across these organizations."

# Filter for the last 5 years
last_5_years <- landings_filtered %>%
  filter(year >= max(year, na.rm = TRUE) - 4) %>%  # last 5 years including max year

  group_by(source) %>%
  summarize(
    avg_pounds = mean(pounds, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(avg_pounds))

# Plot
ggplot(last_5_years, aes(x = reorder(source, avg_pounds), y = avg_pounds, fill = source)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(
    title = "Average Shark Landings by Data Source in the Last 5 Years",
    x = "Data Source (Fisheries Management Organization)",
    y = "Average Pounds Landed",
    caption = "Source: NOAA Fisheries landings data"
  ) +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal()

```

